---
title: "Activity Monitoring Data"
author: "Tony Li"
date: "November 19, 2016"
output: html_document
---

Loading the raw data and appropriate packages into R
====================================================

```{r, message=FALSE}
library(dplyr)
library(readr)
library(ggplot2)

activity <- read_csv("activity.csv")
```

What is the mean total number of steps taken per day?
=====================================================

Summarizing the steps in *activity* data table for each day:

```{r}
activity2 <- activity %>%
    group_by(date) %>%
    summarise(steps = sum(steps, na.rm=TRUE))
```

Here's a sample from *activity2* data table:
```{r echo=FALSE}
activity2[1:10,]
```

Here's a histogram of total steps per day:
```{r}
ggplot(data=activity2, aes(x=steps)) +
    geom_histogram(color="black", fill="green", bins=50) +
    labs(title="Histogram of Total Steps per Day", x="Steps", y="Frequency")
```

We want to find out the average and median daily total number of steps.
```{r mean_med, include=TRUE}
avg_steps_day <- mean(activity2$steps, na.rm = TRUE)
med_steps_day <- median(activity2$steps, na.rm = TRUE)

avg_steps_day
med_steps_day
```

The average daily total number of steps is approximately `r round(avg_steps_day)` steps.  
The median daily total number of steps is `r med_steps_day` steps.

What is the average daily activity pattern?
===========================================

We want to make a time-series plot of the 5-minute interval and the average number
of steps taken, averaged across all days. 

First, summarizing the steps in *activity* data table by interval:
```{r}
activity3 <- activity %>%
    group_by(interval=interval) %>%
    summarise(avg_steps = mean(steps, na.rm=TRUE))
```

Here's a sample from *activity3* data table:
```{r echo=FALSE}
activity3[1:10,]
```


Here's the time-series plot showing the total steps over the 5-min intervals:

```{r}
ggplot(data=activity3, aes(x=interval, y=avg_steps)) +
    geom_line() +
    labs(title="Average Steps over 5-min Interval", x="Interval", y="Steps") +
    scale_x_continuous(breaks=seq(0,max(activity3$interval)+100,200))
```

We want to find out which interval has the maximum number of steps:
```{r include=TRUE}
max_steps_5min <-  activity3[which.max(activity3$avg_steps),]
max_int <- max_steps_5min$interval
max_steps <- max_steps_5min$avg_steps

max_int
max_steps
```
The 5-min interval that contains the maximum number of steps is `r max_int` to 
`r max_int+5` minute which has on average `r round(max_steps)` number of steps.  


Imputing Missing Values
=======================

So far we have been ignoring the existence of missing values ("NA"), which can
introduce bias into some calculations.  

Calcuting the total number of missing values in the dataset:
```{r}
sum(is.na(activity$steps))
```

There are `r sum(is.na(activity$steps))` rows with missing value. Since we already
calculated the average number of steps for each 5-min interval, let's fill in
the missing values with the averages for the interval they are in. Remember that
the averages are in *activity3* data table.

```{r}
idx <- is.na(activity$steps)
na_intervals <-activity$interval[idx]
activity$steps[idx] <- activity3$avg_steps[match(na_intervals, activity3$interval)]
```

Now let's check if there is any missing values in the *activity* dataset:
```{r}
sum(is.na(activity$steps))
```
No more missing values!


Using the new data table, let's plot the histogram again and re-calculate the
mean and median daily total number of steps.  

Here is the histogram:
```{r}
activity2_adjNA <- activity %>%
    group_by(date) %>%
    summarise(steps = sum(steps))

ggplot(data=activity2_adjNA, aes(x=steps)) +
    geom_histogram(color="black", fill="green", bins=50) +
    labs(title="Histogram of Total Steps per Day (Adjusted NA)", x="Steps", y="Frequency")
```

Comparing to the first histogram we plotted without filling in the missing values, we
see that the frequency at 0 steps decreases, and the frequency at center increased.
This is expected since we did fill in the missing data with the average!  

Here's the new mean and median, respectively:
```{r}
avg_steps_day_adjNA <- mean(activity2_adjNA$steps)
med_steps_day_adjNA <- median(activity2_adjNA$steps)

avg_steps_day_adjNA
med_steps_day_adjNA
```

The mean increased from `r as.integer(avg_steps_day)` steps to `r as.integer(avg_steps_day_adjNA)` steps.  
The median increased from `r med_steps_day` steps to `r as.integer(med_steps_day_adjNA)` steps.  
The new mean and median are the same due to the fact that we added the mean to the 
missing values.


Are there differences in activity patterns between weekdays and weekends?
==========================================================================

We will create a new factor variable called *wk* with two levels - "weekday"
and "weekend" indicating whether a given date is a weekday or weekend.

```{r}
activity$wk <- weekdays(activity$date)
activity$wk[activity$wk=="Saturday" | activity$wk=="Sunday"] <- "Weekend"
activity$wk[activity$wk!="Weekend"] <- "Weekday"
```

Here is a sample of the data set with the new variable:
```{r echo=FALSE}
activity[1:10,]
```

Here is a time-series of total steps by 5-min interval, but broken up by weekdays
and weekends:
```{r}
activity3_adjNA <- activity %>%
    group_by(interval=interval, wk=wk) %>%
    summarise(avg_steps = mean(steps))

ggplot(data=activity3_adjNA, aes(x=interval, y=avg_steps)) +
    geom_line() +
    facet_grid(wk~.) +
    labs(title="Average Steps over 5-min Interval", x="Interval", y="Steps") +
    scale_x_continuous(breaks=seq(0,max(activity3_adjNA$interval)+100,200))
```





















